### 2025-04-08 Paper.tex
update Paper training

@@

\subsection{Manual Augmentation}

@@

@@

\color{red}INSERT OTHER CRITERIA HERE\color{black}

>>>>

	%==============================================
	%================ Model Architecture =================
	%==============================================
	\section{Model Architecture}

	\subsection{Different approaches in VC}

	To generate recent tracks, but with the voice of Chester Bennington, we must introduce a method to convert or synthesize singing voices. Therefore, we briefly introduce the field of voice conversion, focussing on singing voice and singing voice synthesis. Voice conversion (VC) describes the transformation from a source speaker identity to a different speaker identity, while preserving the articulated linguistic content. During this process, different parameters of the voice, such as affect, emotion and accents are manipulated, to improve the naturalness and the similarity to the target voice. General breakthroughs that facilitated the study of voice conversion are the debut of computer based speech synthesis in the 1950s, the progress in digital signal processing during the 1970s and the recent advances in Machine Learning, especially induced by Neural Networks. \cite{Sisman2020} As a result, the difficulties of voice conversion have changed from solely generating natural sounding voices to more advanced applications, for example creating voices from scratch, that are not based on human training data.

	Singing Voice Conversion (SVC) is considered a special case of VC, dealing with the conversion of sung voices. Additionally to the linguistic content, we want the pitch to be preserved as much as possible and focus more on the naturalness of the generated target voice (in terms of intonation, sound quality, timbre, energy and singing style). SVC faces similar challenges and difficulties as standard voice conversion but relies on further style parameters, which have to be trained. We contrast between parallel and nonparallel singing voice conversion: parallel SVC refers to models, that do a one-to-one conversion (which means they are only capable to convert the voice pair they trained on), while nonparallel SVC refers to any-to-one or any-to-any systems, where the target speakers are decoded via an additional input speaker-ID. There exist multiple well-known SVC models using different approaches, such as SOVITS which is combining Soft-VC and VITS, RVC which consists of an additional retrieval part and DDSP-SVC which uses a shallow diffusion mechanism. \cite{MediumSVC2023} A major problem SVC faces is data sparsity - due to copyright reasons we only have few published singing databases, often consisting of traditional folk songs.

	The less naive approach to generate a target voice is singing voice conversion (SVS). Singing voice synthesis models take the lyrics, the musical score and the respective duration as input and generate the audio file from scratch. During the training process the model predicts frame-level information of the target voice, which are then transferred to an audio signal by a vocoder. \cite{Cho2021} A prominent SVS system is the DiffSinger architecture.

	The singing voice conversion and the singing voice synthesis approach both come with advantages and difficulties, on which our model choice is based. SVC is a manipulation of the source audio and therefore often outperforms SVS models in preserving the singers expressiveness and naturalness, whereas SVS grant the user full control over the model and much more flexibility regarding the input songs (because it’s not reliant on existing performances). More flexibility increases the data and time consumption of synthesis models during training, which leads to higher computational costs compared with conversion models.

	Another significant factor was the public data availability for our project. We have plenty of training data for our target voice (Chester Bennington), including 7 studio albums, 3 live albums and even more live performances - but we encounter a sparse data situation for Emily’s voice, where only a single album is published and even less audios are available, where both perform the same song.

	Regarding all advantages and factors mentioned above, we concluded that singing voice synthesis models fit our project better than the naive conversion approach. The sparse data situation for songs jointly performed by target and source singer, which is substantial for SVC training, forces us to focus on synthesizing methods. Furthermore we gain more flexibility and control over the model, which enables us to be more creative with the input data and gives light to a potential follow-up project - to generate new lyrics and songs with the voice of Chester Bennington.

	\subsection{Singing Voice Synthesis (SVS)}

	The breakthroughs in Deep Learning enabled novel possibilities for the generation of singing voices and in the past decade a variety of modern SVS architectures were proposed. While the majority of models are adapted voice conversion systems with additional singing-specific features, some of the proposed architectures introduce completely new concepts to Machine Learning and partly even influenced the field of voice conversion. Research focuses especially on models improving the naturalness of the output voice and models developed for low data situations. Subsequently, we travel through some of the most influential SVS models from the past decade to make our decisions transparent and comprehensible. \cite{Cho2021}

	Early systems are often based on hidden markov models (HMM), which consist of a rather plain architecture and comparatively simple to interpret. Consequently, the generated results are outdated nowadays and cannot reach the standard of recent models. \cite{Cho2021}

	The vast majority of recent SVS models use deep neural networks (DNNs) to preserve natural and expressive outputs. Important is the partition into submodules: an acoustic model, that is trained to predict acoustic and pitch related features and a neural vocoder that generates the audiowave based on the predicted information. This aspect of divide and conquer plays a decisive role in the success of neural models. Well-known DNN architectures are XiaoiceSing \cite{Cho2021}, which adopts the FastSpeech text to speech model and extends it with note duration and pitch information prediction; HifiSinger \cite{Cho2021}, a model generally based on XiaoiceSing, but using a GAN vocoder to prevent the model from over-smoothing; DiffSinger \cite{DiffSinger2021}, which improves the weaknesses of the previous models by introducing a diffusion probabilistic model (= gradually applies noise to the mel-spectrogram until the noise is gaussian distributed with subsequent denoising steps back to a mel-spectrogram).

	Apart from different approaches to enhance the target voice naturalness, low resource availability is a prevailing challenge in singing voice synthesis. Numerous models were designed to be more data efficient, requiring only small portions of training data, compared to the conventional models. The Sinsy \cite{Cho2021} model produces proper results with 1 hour of conditioned training data, partly due to a partition of the models architecture into further submodules. Zero shot models (for example StyleSinger and YourTTS) request merely minutes of training data to transfer voices in different languages or styles. \cite{Casanova2021},\cite{Zhang2024}

	Expressive and powerful models are also increasingly attracting the interest of large tech companies to market voice generators commercially. In 2022, Microsoft published its software DeepSinger \cite{Ren2020}, which conducts data mining on music websites to generate more training data.

	\subsection{Model choice}

	Based on the presented architectures, we aimed to choose the model fitting our requirements the best. The model should produce natural audios and be capable of capturing the expressiveness of Chester Bennington’s voice. Further it should be comparatively easy to reimplement and well documented for research purposes. Data sparsity has solely been an issue for the source singer's voice, since Chester’s voice is published on numerous albums and publicly available.

	Summarizing all models, we are of the opinion that the Diffsinger model meets our expectations the best and makes full use of the possibilities available to us, because it generates high-quality results and is widely documented with an active community of (non-commercial) users. Furthermore we are not reliant on computational and data limitations for our experiments and can therefore exploit the full potential of complex neural networks.

@@

	\subsection{Data Augmentation}

	Data augmentation describes the artificially increasing of the available training data to stabilize the training process, improve the performance and to diversify small and incomplete data sets. For this purpose, the existing training data is slightly modified and manipulated, reducing the  time and cost consumption that is often needed to generate real data. Previous studies showed that in voice conversion carefully designed augmentation is crucial. In particular additive noise and extensive data manipulation resulted in significant performance decreases and solely minor data changes achieved comparable accuracy. \cite{Slizovskaia2022}  For SVS, research groups proposed several different data augmentation methods with limited success. Experiment setups including cycle-consistent data augmentation or varying segment durations gained favorable results. \cite{Guo2022},\cite{Zhang2022} Also pitch shifting with a restricted range showed slight improvement of SVS models. \cite{Guo2022} Yet, none of the methods indicated a significant performance boost that outperformed all other methods.

	Therefore, we augmented the data with two different approaches, first using the built-in application of the GitHub repository and second performing manual pitch and speed transformations.

	\subsubsection{Built-in Configuration}

	The openvpi DiffSinger repository allows two different ways of augmenting data, which can be controlled in the acoustic model’s configuration file. With the knowledge to augment data cautiously, we retained the preset parameters of the model. DiffSinger supports either fixed or random pitch shifting (default variant is fixed) ranging from [-5, 5] semitone divergence to the original data and time stretching in the range from half speed to double speed. \cite{openvpi-diffsinger}

	\subsubsection{Manual Augmentation}

@@

	\subsection{Training Setup}

	The training setup and training organisation might seem less interesting and relevant at first, but the entire process (including the server usage) was particularly instructive for us and therefore deserves a section in the documentation. Opposing to Neil Armstrong - “That’s one small step for mankind, one giant leap for us” .

	The openvpi DiffSinger repository \cite{openvpi-diffsinger} offers significant advantages over the DiffSinger version that was implemented for the respective paper publication \cite{moon-diffsinger}; sufficient documentation is available, openvpi comes with an integrated variance module and the code readability strongly improved. The latter is particularly helpful, when we face problems that can only be solved by debugging.

	Initially we based our training setup on existing Google colab notebooks and other publicly available tutorials and used the procedures as orientation. Though  with colab notebooks, we encountered some fundamental and major problems relatively quickly, which forced us to build our setup from scratch. To avoid GPU usage limitations, we trained all our models on the LST servers of the Saarland University. \cite{lst-wiki}

	The LST cluster of the Saarland University uses HTCondor to schedule jobs and distributes them to free GPUs in incoming order. Users request the necessary computational cost for a job via submit files and transfer the file to be executed (must be a bash script). The dependencies are managed inside the bash script with an anaconda environment or in the submit file using a docker. For this purpose, we created an anaconda environment running on python version 3.8.20 and installed the necessary requirements following the DiffSinger installation guide. \cite{openvpi-diffsinger}

	Though HTCondor provides the possibility to distribute the executed job among several GPUs to decrease the running time, we restricted our training to 1 GPU (less potential to run into errors) and requested 16GB of GPU memory. After running for a maximum of 1 day, jobs get automatically kicked from the server by default. The image shows an example submit file that we used in our project.

	\color{red}image of example submit file\color{black}

	To gain control for optimal stopping points, we monitored the training process with tensorboard, which visualizes the current epoch, the course of training loss and validations loss and the results validation samples.

	\subsubsection{Problems}

	The configuration of the training setup caused us many problems for any number of reasons, which made the process an exhausting task. Often, installation and training guides from the internet were outdated and used DiffSinger versions that are no longer supported. Then for the first trial on our setup, we used Docker, which ultimately failed to search the python modules in the right folder, forcing us to find an alternative (anaconda environment). And once finished with the setup, trivias like a missing LST cluster specification delayed our project permanently.

	\subsection{Training}

	We trained nine different versions of the DiffSinger model, applying various configuration alternations and providing datasets of different size and audio quality. All models were trained under equal conditions, which means that we always used the same conda environment for the runs and submitted, up to minor changes, identical bash scripts. The majority of training parameters remained constant and unchanged throughout the entire training procedure - therefore the configuration parameters in question are presented once at the beginning and apply to all training trials. Modified values (for example data augmentation) and integrated modules (for example the usage of a variance model) get specified in the section of the particular run.

	The main architecture of the DiffSinger model is trained to predict mel-spectrograms out of the input data, but that is not exactly what we challenged ourselves. To generate audio files from the mel-spectrograms a vocoder is necessary. For a start we stuck to the HifiGAN vocoder recommended in the openvpi DiffSinger support, which we ultimately fine-tuned on Chester’s voice (training see below in the section vocoder).

	The repository offers several built-in models operating as pitch extractor and harmonic-noise separator as well, recommending RMVPE for extracting pitch and Vocal Remover (VR) to separate the harmonic and aperiodic parts of the singer's voice. For the shallow diffusion configuration again we were oriented strongly towards the given guide, copying the preset parameter values.

	Another relevant decision, to minimize the training loss and presumably improve the mel-spectrograms, is the right choice of optimizer and learning rate. Previous experiments and the paper itself achieved promising results with the AdamW optimizer and different learning step sizes. So there was no reason for us to integrate different optimizers than AdamW and unless otherwise stated we set the learning rate to 0.0006.

	To preserve high-quality audio outputs we used as sample rate 44.1 Hz, with a hop size of 512 (which is the step length of mel and feature extraction) and a balance between frequency resolution and efficiency during the mel-spectrogram extraction, setting 2048 for the Fast Fourier Transform size.

	\subsubsection{Vocoder}

	Another tool to enhance the overall performance, is fine tuning the vocoder on the target voice. This should in theory improve the transformation process from mel-spectrogram to audio, in particular preserving the singer identity characteristics and the singing style. For this purpose, we trained the recommended HifiGAN model checkpoint on the cleaned (hand segmented) dataset for a few epochs to personalize the vocoder. The AdamW optimizer updated the vocoder parameters with learning rate 0.0001 and batch size 6. Different to the DiffSinger model, the HifiGAN vocoder was fine-tuned using parselmouth as pitch extractor. During training we observed convergence of the validation loss relatively quick after around 2000 epochs.

	\subsubsection{Model 1 (small)}

	The first generated model served primarily as a test trial to check whether the entire workflow, from extracting input data to generate outputs, functions. Unsurprisingly we kept the standard configuration from above, without using any form of data augmentation or fine tuning and without using a pretrained variance model. The dataset consisted of all available studio albums and we used 6 segments for validation and we used a batch size of 64. We stopped the training after roughly an hour and 9900 trained epochs, while the training loss was still on decline - though the validation loss already stabilized around 0.04 (as displayed in image \color{red}v1\color{black}). In the following we will refer to this model as small.

	\color{red}image of v1 (training+testing)\color{black}

	\subsubsection{Model 2 (small)}

	The second generated model serves as an extensive version of the small model in the section above. All configuration and hyperparameters were transmitted with the only difference of a longer training duration until the training loss stabilizes. After a little more than 70k epochs, the aux mel training loss decreased to 0.009 and the mel training loss reached the minimum of 0.06. Worth mentioning is the increasing fluctuation of the validation loss, starting after reaching the minimal value (see image \color{red}v2\color{black} ). Due to no subjective improvement of the model compared to Model 1, this version is not sustained in the following evaluation process.

	\color{red}images of v2 (training, testing)\color{black}

	\subsubsection{Model 3 (large)}

	The idea for the third model was to enlarge the training data size to observe the effect on the model’s performance. So, we extended the data to all available Linkin Parks albums, studio albums as well as live albums but without applying any additional data augmentation techniques. We kept the batch size of 64, but increased the amount of validation samples (now 10) to balance the training validation ratio. The enlarged dataset decelerated the training process massively, reaching 16 hours and 78k epochs of training until the training loss converged. Final loss values comparable to those of the small model were reached during training, minimal value of 0.01 for the training aux mel loss and 0.028 for the validation aux mel loss (see image \color{red}v3\color{black} ). This model is referred to as large during evaluation.

	\color{red}image of v3 (training+testing)\color{black}

	\subsubsection{Model 4 (clean)}

	Next, we generated a clean model, to investigate the relevance of noisy input data on the model performance and audio quality of the results. Therefore we trained the model solely with the reduced hand segmented dataset, that does not contain noisy data or segments with Mike Shinoda’s voice. Due to the smaller dataset we reduced the amount of segments per batch to 64, trained with a lower learning rate of 0.004 and used only 8 samples for validation. Subsequently the duration until loss conversion shrunk to 52k epochs and 6 hours training time. As displayed in image \color{red}v4\color{black}, the training mel loss reached a minimum of 0.006 and the validation mel loss of 0.03. This model will be referred to as clean.

	\color{red}image of v4 (training+testing)\color{black}

	\subsubsection{Model 5 (large+ft+vocoder)}

	After probing the relevance of dataset size and data quality, we investigated whether fine-tuning with clean data shows a positive effect. To receive encompassing trends, the fine tuning is also extended to the vocoder. We pretrained the model with the exact same configurations as the large model, though integrated the fine-tuned HifiGAN vocoder instead of the standard vocoder. Following up on the pretraining, where much but low-quality data was used, the model is fine-tuned on the clean dataset to focus on learning the subtle nuances in Chester’s voice. The validation set for the fine-tuning consists of 8 validation samples, all other parameter values were retained from the pretraining. During training, the model confirmed 8000 epochs and ran about 1.5 hours before the validation loss reached convergence. The aux mel training loss minimized to 0.007 and the aux mel validation loss to 0.035 (see image \color{red}v5\color{black} ). This model will be referred to as ft+vocoder.

	\color{red}image of v5 (training+testing)\color{black}

	\subsubsection{Model 6 (huge+variance+vocoder)}

	The task of the variance+vocoder model was to investigate whether the combination of variance and acoustic model is superior over the standard acoustic model and which importance the pitch and tension prediction takes for our experiment. The process is parted into two steps; first we trained the variance model on the large dataset for 11 hours and 97k epochs, then we inferred the large dataset with the variance model and trained the acoustic model on them for 2.5 hours and 17k epochs. The variance model was configured to predict pitch and tension of Chester’s voice. Additionally to the large dataset we further enriched the variance data with the DiffSinger built-in data augmentation method for the acoustic model training.  For both models we integrated the fine-tuned vocoder, used batch size 64  and 10 sample segments for validation. When stopped training the acoustic model achieved a mel aux loss of 0.017 for training and 0.028 for validation (see image \color{red}v6\color{black} ), further the variance model reached a pitch accuracy of 0.5.

	\color{red}image of v6 (training+testing)\color{black}

@@

	\subsubsection{MOS prediction}

	A basic but quite expressive evaluation metric is the mean opinion score (MOS), which originates in the research of telecommunication to assess the quality of human speech. The metric is standardized and quantitative, ranging from values between 0 (worst quality) and 5 (best quality) - where 0 means, communication is not possible despite high efforts and 5 means, communication is afforded effortlessly.

	Previous work, adapted the MOS for voice conversion models, generating automatic mean opinion score prediction models that are now broadly used for voice conversion evaluation. Further applications for automated MOS are for example the detection of artificially generated fake audios or the evaluation of other assessment tools as gold standard. \cite{Zhou2024} However, in the past automated MOS prediction primarily served as a tool to compare different models and model architectures and not for hyperparameter or database testing. Nevertheless, we regard it as a solid basis for further evaluation metrics.

	For our project, we integrated the SingMOS model \cite{Tang2024}, a predictor particularly developed for the evaluation of singing voice synthesis models. The collected test dataset consist of 15 segments, purposed to provide a broad coverage of Chester’s different singing styles including screams and shoutings (all segments are as noiseless as possible). The finetune+vocoder and variance+vocoder model were unable to generate inferences of one test segmented, so we left the affected segment out for both models. Additionally we predicted the MOS for the vocal separated audio files to receive a gold standard value - scoring a value of 2.995.

	\color{red}table of mos results\color{black}

	All results are averaged across the models and summarized in \color{red}table 1\color{black}. As expected none of the models surpasses the ground truth value of the vocal separated segments. Additionally the distance between the highest and lowest performing model is with 0.22 score points difference only marginal.

	The variance+vocoder model outperformed all other models and reached results comparable to the ground truth, indicating outputs with relatively high audio quality. At some distance follows the result of the manual+clean model, achieving a slightly worse but still comparably high predicted MOS score. What is striking is that in particular models trained on smaller datasets (independent of audio quality) seem to receive lower MOS prediction values. We suspect this trend to be provoked by noise in the output audios.

	\subsubsection{Histogram Comparison}

	In the next step, we wanted to evaluate the models based on a comparison of the mel spectrograms. A recent blog post introduced a method, where the mel spectrograms are converted to histograms (distribution of the bin values), which are easier to compare. \cite{mel-spectrogram} This approach resembles the comparison of the underlying bin distribution and therefore takes the place of a meta evaluation.

	To measure the difference of the generated histograms, we applied several common comparison metrics. First measurement is the Correlation of both histograms, with values ranging from -1 (totally uncorrelated) to 1 (totally correlated) - which draws an extremely direct connection between both bin distributions. Another popular measurement to compare histograms and distributions in general is the Chi-Square distance. Strength of the metric is the sensitivity for large distances, meaning distributions that deviate strongly from the original histogram can be detected easily and filtered out. Obtained values range from 0 to infinity, the larger the value the more mismatches the distribution. A completely different perspective allows the Intersection measure of the histograms, where the overlapping of equal bin values is computed. With the comparatively low complexity and computational cost comes quite inaccurate results, that might be fluctuating and less expressive for model performance. The higher the Intersection value, the more bin values share two distributions. A more refined measurement is the Bhattacharyya distance, which originates in the measurement of probability distribution similarity. Similar to the Chi-Square distance, the smaller the value, the higher is the distribution’s similarity.

	\color{red}table of results histogram comparison\color{black}

	We implemented all metrics quoted above and tested them on the entirety of our models, using the same test set as for the MOS prediction. Each model is compared with the vocal separated data as our gold standard. The obtained results are averaged for each model and summarized in \color{red}table 2\color{black}.

	The variance+vocoder model outperforms all other models, achieving constant and strong results in almost every comparison measurement, indicating a clear dominance between the tested models. 
	For the Correlation values we observe just minor differences, but the variance+vocoder model and the large model achieve the highest results with scores around 0.95.

	While keeping in mind that the Intersection value might possibly be less informative than other comparison measurements, the variance+vocoder and large model again achieve the highest results, with the most overlapping of bin values. Furthermore, models trained or fine-tuned on the clean dataset perform the worst with this metric. Potentially, the “high quality” data is shifting or skewing the results in either direction, which preserves general distribution properties but changes the bin overlap.

	The Chi-Square distance remains the only measurement, where the variance+vocoder model is beaten by other models - with the small and large model outperforming the rest, both reaching values below 300000. However, no enlightening tendencies or correlations between model hyperparameters and Chi-square distance are observable, therefore the values appear to be rather meaningless for the evaluation.
	The variance+vocoder model scores the smallest Bhattacharyya distance with the value 0.16, once more outperforming all other tested models. In general models with the fine-tuned vocoder reached comparatively good results, where further research is necessary to investigate potential reasons for this effect.
